\documentclass[11pt]{report}

% Document dimensions
\usepackage{geometry}
\geometry{top=1.5cm, bottom=1.5cm, textwidth=15cm}

% Math related packges.
\usepackage{amsmath}
\usepackage{cancel}

% Natural Deduction package
\usepackage{proof}

% Fix the header space: start at the top of the page.
\usepackage{hyperref}

\begin{document}
	
% Heading for the tutorial	
\begin{center}
	{\bf MATH230: Tutorial Four}
\end{center}
\begin{center}
	{\bf Test Revision}
\end{center}

% Box with goals and relevent lecture notes.
\noindent\fbox{
	\parbox{\textwidth}{

	Key ideas:
		\begin{itemize}
			\item Revise the propositional logic topic.
		\end{itemize}

	Relevant content: Lectures 1 - 13 and Tutorials 1,2,3\\
	Relevant reading: \href{https://leanprover.github.io/logic_and_proof/index.html}{L$\exists\forall$N Chapters 1,2,3,5,6}  
		
	\vspace{0.2cm}

	Hand in exercises: {\bf 2022 test}
	}
}
% Discussion questions for tutor.
\newline
\vspace{0.5cm}

\noindent {\bf Discussion Questions}

\begin{enumerate}
	% Use valuation to show not semantic consequence.
	\item Use valuations to decide whether the following propositions are tautologies, contingent, or contradictions.
		\begin{enumerate}
			\item $(\lnot P \land \lnot Q) \ \to \ \lnot(P \land Q)$ % Tautology
			\item $\lnot(P \land Q) \ \to \ (\lnot P \land \lnot Q)$ % Contingent
			\item  $\lnot(P \lor Q) \to (P \land Q)$% Contradiction
		\end{enumerate}
	
	\vspace{4cm}

	% Use RAA in natural deduction.	
	\item Provide a natural deduction to prove $\vdash \ A \lor \lnot A$.
	
	\vspace{4cm}

	% Use Soundness to conclude no derivations possible.
	\item Explain why there can be no derivations of the proposition $A \land \lnot A$. Your answer should reference a metalogical theorem.
\end{enumerate}

% New page for tutorial exercises.
\newpage
{\bf Tutorial Exercises}
\begin{enumerate}
	
	\item Work through the 2022 midterm test. 
	
	\item Prove each of the following claims of semantic consequence 

		\begin{enumerate}
			\item $\lnot P \ \land \lnot Q \ \models \ \lnot(P \lor Q)$
			\item $\lnot(P \lor Q) \ \models \ \lnot P \ \land \lnot Q$
			\item $\lnot P \ \lor \lnot Q \ \models \ \lnot(P \land Q)$
			\item $\lnot(P \land Q) \ \models \ \lnot P \ \lor \lnot Q$
		\end{enumerate}

	\item Explain how we know there must be natural deductions proving each of the corresponding syntactic consequences. Your answer should appeal to an appropriate metalogical theorem. 

	\item Provide natural deductions for the following claims of syntactic consequence 

		\begin{enumerate}
			\item $\lnot P \ \land \lnot Q \ \vdash \ \lnot(P \lor Q)$
			\item $\lnot(P \lor Q) \ \vdash \ \lnot P \ \land \lnot Q$
			\item $\lnot P \ \lor \lnot Q \ \vdash \ \lnot(P \land Q)$
			\item $\lnot(P \land Q) \ \vdash \ \lnot P \ \lor \lnot Q$
		\end{enumerate}
	
	One of these natural deductions will \emph{require} the use of RAA. Use the BHK interpretation of the propositional connectives to figure out which one.

	\item Prove the following claim twice:
	
		$$ \cancel{\models} \ A \to \lnot A$$

		\begin{enumerate}
			\item with an argument using a truth table, and
			\item with an argument using a valuation function.
		\end{enumerate}
		
	

\end{enumerate}
	
\end{document}