\documentclass[11pt]{report}

% Document dimensions
\usepackage{geometry}
\geometry{top=1.5cm, bottom=1.5cm, textwidth=15cm}

% Import the necessary preamble for the document. 
\usepackage{../../../proofsPrograms}

% Math related packges.
\usepackage{amsmath}
\usepackage{cancel}

% Natural Deduction package
\usepackage{proof}
\usepackage{mdframed}

% Fix the header space: start at the top of the page.
\usepackage{hyperref}


\begin{document}
	
	
% Heading for the tutorial	
\begin{center}
	{\bf MATH230: Tutorial Seven}
\end{center}
\begin{center}
	{\bf Peano Arithmetic}
\end{center}


% Box with goals and relevent lecture notes.
\noindent\fbox{
	\parbox{\textwidth}{

		Key ideas
			\begin{itemize}
				\item Natural deduction practice, 
				\item Proofs using the identity rules of inference,
				\item Prove first-order sentences in theories of arithmetic,
				\item Use the induction schema of Peano arithmetic, and 
				\item Become exasperated enough to appreciate the help of proof assistants. 
			\end{itemize}

		Relevant lectures: Lectures 14,15, and 16\\
		Relevant reading: \href{https://www.ma.imperial.ac.uk/~buzzard/xena/natural_number_game/}{Natural Number Game}
		
	\vspace{0.2cm}

	Hand in exercises: 1a, 2a, 2b, 2c, 3a, 3b, 3c\\ 
	{\bf Due following Friday @ 5pm to the tutor, or to lecturer.}
	}
}
% Discussion questions for tutor.
\newline
\vspace{0.5cm}

\noindent {\bf Discussion Questions}

	Proofs that make use of the axiom (schema) PA 7 

	$$[P(0) \land \forall x \ (P(x) \to P(s(x)))] \rightarrow \forall y (P(y))$$

	will prove statements of the form $\forall y \ P(y)$ with the use of modus ponens. This requires proving the antecedent conjunction: 

	$$[P(0) \land \forall x \ (P(x) \to P(s(x)))]$$

	This in turn requires proving each conjunct i.e. two proofs witnessing: 

	$$\PA \vdash P(0) \hspace{2cm} \PA \ \vdash \ \forall x \ (P(x) \to P(s(x)))$$

	If we piece this together then we see that all proofs by induction have the form: 

	\begin{mdframed}
		\begin{center}
			$\begin{array}{c}
				\infer[\IND]{\forall y \ P(y)}
					{\infer[\land I]{[P(0) \land \forall x \ (P(x) \to P(s(x)))]}
						{\infer[]{P(0)}{\begin{array}{c} \vdots \\ \mathcal{D}_{BC}\end{array}}
						&
						\infer[\forall I]{\forall x \ (P(x) \to P(s(x)))}
							{\begin{array}{c} \vdots \\ \mathcal{D}_{IS}\end{array}}}}
			\end{array}$
		\end{center}
	\end{mdframed}

	\begin{enumerate}
		\item Identify the following steps involved in a proof by induction of the following sentence of Peano arithmetic: 
		
		$$\PA \ \vdash \ \forall x \ (0+x=x) $$

			\begin{enumerate}
				\item Identify the wff $P(x)$ to do induction on. 
				\item $\mathcal{D}_{BC}$: Write down the sequent $\PA \ \vdash \ P(0)$.
				\item $\mathcal{D}_{IS}$: Write down the sequent $\PA, P(n) \ \vdash \ P(s(n))$.	
			\end{enumerate}
	\end{enumerate}

% New page for tutorial exercises.
\newpage
{\bf Tutorial Exercises}
\begin{enumerate}

	\item Give natural deductions of the following theorems of identity.
	
	\begin{enumerate}
		\item $\vdash \ \forall x \forall y \forall z \ (x = y \land y = z) \to x = z$
		\item $\vdash \ \forall x \forall y \forall z \ x \neq y \to (x \neq z \lor y \neq z)$ \hspace{5cm} (RAA)
		\item $\vdash \ \exists x \ (t = x)$
	\end{enumerate}

	Part (a) together with proofs from lectures imply that identity is reflexive, symmetric, and transitive. Thus behaving like an equivalence relation. 
	
	\item In this question $\text{PA}$ denotes the first-order theory of Peano arithmetic which has signature PA$: \{0,s,+,\times\}$ and axioms:
	
	\begin{enumerate}
		\item[PA1] $\forall x \lnot(s(x) = 0)$
		\item[PA2] $\forall x \ \forall y ((s(x) = s(y)) \to (x = y))$
		\item[PA3] $\forall x \ (x + 0 = x)$
		\item[PA4] $\forall x \ \forall y \ (x + s(y) = s(x + y))$
		\item[PA5] $\forall x \ (x \times 0 = 0)$
		\item[PA6] $\forall x \ \forall y \ (x \times s(y) = (x \times y) + x)$
		\item[PA7] $[P(0) \land \forall x \ (P(x) \to P(s(x)))] \rightarrow \forall y (P(y))$
	\end{enumerate}	
	
	Provide deductions to prove the following sequents. 
	
	\begin{enumerate}
		\item $\text{PA} \vdash 1 + 1 = 2$
		\item $\text{PA} \vdash 3 \neq 1$
		\item $\text{PA} \vdash \forall x \ (x + 1 = s(x))$
		\item $\text{PA} \vdash \forall x \ (x \times 1 = x)$ 		
	\end{enumerate}

	\newpage
	\item The followings sequents all require the use of the induction axiom schema. Recall that all proofs using the induction schema have the following form: 
	
		\begin{mdframed}
			\begin{center}
				$\begin{array}{c}
					\infer[IND]{\forall y \ P(y)}
						{\infer[\land I]{[P(0) \land \forall x \ (P(x) \to P(s(x)))]}
							{\infer[]{P(0)}{\begin{array}{c} \vdots \\ \mathcal{D}_{BC}\end{array}}
							&
							\infer[]{\forall x \ (P(x) \to P(s(x)))}
								{\begin{array}{c} \vdots \\ \mathcal{D}_{IS}\end{array}}}}
				\end{array}$
			\end{center}
		\end{mdframed}

	For this reason, once the wff $P(x)$ is identified, it suffices to provide the base case deduction $\mathcal{D}_{BC}$ and induction step $\mathcal{D}_{IS}$. The sequents are stated in such a way as to mean induction on the variable $x$ will be the easiest approach. Always do induction on the variable $x$.  
	
	\begin{enumerate}
		\item $\text{PA} \vdash \forall x \ (0 + x = x)$
		\item $\text{PA} \vdash \forall x \ (0 \times x = 0)$
		\item $\text{PA} \vdash \forall x \ (1 \times x = x)$
		\item $\text{PA} \vdash \forall x \ (x = 0 \lor \exists y (x = s(y)))$ \hspace{5.2cm} (Challenge!)
		\item $\text{PA} \vdash \forall x \ \forall y \ [s(y) + x = s(y+x)]$ \hspace{5cm} (Challenge!)
		\item $\text{PA} \vdash \forall x \ \forall y \ \forall z \ [(y + z) + x = y + (z + x)]$ \hspace{3.5cm} (Challenge!)
		\item $\text{PA} \vdash \forall x \ \forall y \ [y + x = x + y]$ \hspace{5.98cm} (Challenge!)
	\end{enumerate}

	\item Visit the \href{https://www.ma.imperial.ac.uk/~buzzard/xena/natural_number_game/}{Natural Number Game} to write computer checked proofs of these theorems of Peano arithmetic as well as theorems of propositional logic from previous tutorials. 
	
	\item The first-order language of Peano Arithmetic is often presented with an extra binary relation symbol $<$ where $x < y$ is given the usual interpretation: $x$ is strictly less than $y$. In fact it is not necessary to add anything extra, for this relation can be defined using a sentence in $\PA$ as stated. 
	
	Write down a wff in $\PA$ which, when interpreted in the standard model, defines the binary relation $<$ of being ``strictly less than''. Use this to write down formulae that represent: less than or equal to, strictly greater than, and greater than or equal to. 

	%$$ x < y :\equiv \exists z \ (x + s(z) = y) $$

	\item Write down well-formed formulae in the first-order language of $\PA$ corresponding to the following statements. 
	
	\begin{enumerate}
		\item Each natural number is either equal to $0$ or greater than $0$. 			
		\item If $x$ is not less than $y$, then $x$ equals $y$ or $y$ is less than $x$.
		\item If $x$ is less than or equal to $y$ and $y$ is less than or equal to $x$, then $x=y$. 
	\end{enumerate}
	 
\end{enumerate}	
\end{document}